A first look at the GD-1 stellar stream with Gaia DR2

### 2018-04-28 paper/gd1.tex
intro

@@

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\textsl{PanSTARRS}}
\newcommand{\DR}{\acronym{DR2}}
\newcommand{\msun}{\rm M_\odot}

@@

Dynamically cold stellar streams are formed from the tidal disruption of stellar
systems by the gravitational field of their host galaxy (e.g.,
\citealt{Johnston:1998}).
Over 30 candidate stellar streams have been discovered throughout the halo of
the Milky Way and have a wide range of properties such as length, velocity
dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
recent review \citealt{Grillmair:2006, Newberg:2016}).
Of the thin, likely globular-cluster-origin streams, only two have clear
progenitor systems: NGC 5466 (\citealt{TODO}) and Palomar 5 (\citealt{TODO}).

Stellar streams are

\citealt{Bonaca:2018}

Dynamical modeling of stellar streams or models of mean stream tracks have been used

Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.

We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.

>>>>

% Story:
% - progenitor - show std_v, std_phi2
% - 2 gaps, 1 under-density - epicycles/streakline? percentiles in polynomial
% - spur - encounter?

@@

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\textsl{Pan-STARRS}}
\newcommand{\DR}{\acronym{DR2}}
\newcommand{\msun}{\rm M_\odot}
\newcommand{\article}{\textsl{Letter}}

@@

Dynamically cold stellar streams are formed from the tidal disruption of stellar
systems by the gravitational field of their host galaxy (e.g.,
\citealt{Johnston:1998}).
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution and shape of mass on galactic
scales (\citealt{Bonaca:2018}).
Most stellar streams are found in the halos of galaxies and are therefore
important tracers of the large-scale distribution of dark matter in galactic
halos.

Streams formed from globular cluster systems are particularly useful as they can
theoretically remain coherent for tens of orbital periods, depending in detail
on the symmetries of the underlying galactic mass distribution
(\citealt{Erkal:...}).
This makes them excellent laboratories for studying small-scale structures in
the mass distribution, as it is expected that density perturbations caused by
interactions with massive perturbers can persist and remain visible for close to
their survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
studying or ruling out the existence of small-scale dark matter sub-halos, as
would be predicted by standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal, Sanders, Bovy} papers).

While globular cluster streams are hard to detect around distant galaxies
because of their low relative surface density, over 30 candidate stellar streams
have been discovered throughout the halo of the Milky Way, largely thanks to
large-area, multi-band photometric surveys (e.g., the Sloan Digital Sky Survey,
SDSS; \citealt{York:2000, TODO}).
The Milky Way streams have a wide range of properties such as length, velocity
dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority have no
obvious, bound progenitor systems; Of the thin, likely globular-cluster-origin
streams, only two have clear progenitors (NGC 5466 and Palomar 5; \citealt{TODO,
TODO}).

The most prominent and longest (in angle) thin stream is the GD-1 stream, named
after its discoverers (\citealt{Grillmair:2006}).
GD-1 was found by matched-filtering photometry from the SDSS and spans over
$60^\circ$ in Heliocentric sky coordinates, partially because of its relative
closeness to the sun ($\approx 7$--$10~\textrm{kpc}$).
Despite its proximity, no remnant progenitor cluster or stellar system has been
found associated with the stream.
[Long, somewhat close: useful for constraining large-scale potential (\citealt{Koposov:2009})]

[Small velocity dispersion (quote number?), likely old, retrograde orbit far
from galactic disk, so prime place to search for relics of past subhalo
interactions]
[Known density wiggles and blah (\citealt{DeBoer:2017}), but color-magnitude only, bin stars]

In this \article, we present the cleanest view of the GD-1 stream using
astrometric data from the \gaia\ mission data release 2 (\DR).
Combined with photometry from \pans\ ...
We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.
